---
title: "PA1_template"
author: "SM"
date: "2023-12-08"
output: 
  html_document: 
    keep_md: yes
---

```{r load_packages}
library(tidyverse)
library(knitr)
library(rmarkdown)
```

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE, include=TRUE)
```

Load and process the raw data
```{r load_data}
rawdata<-read.csv("activity.csv", header = TRUE, sep=",")
rawdata$date <- as.Date(rawdata$date, format="%Y-%m-%d")
```

---

Calculate totals, mean, and median number of steps per day
```{r meantotals}
tabdata <-subset(rawdata, is.na(steps)==FALSE,1:2) %>%
  group_by(date) %>%
  summarize(Total_Steps=sum(steps,na.rm=TRUE))
  hist(tabdata$Total_Steps, main="Total Steps Taken Per Day", xlab="Number of Steps", ylab="Frequency")
```

The mean and median number of steps taken per day were **`r format(mean(tabdata$Total_Steps, na.rm=TRUE),scientific=FALSE)`** and **`r median(tabdata$Total_Steps, na.rm=TRUE)`**, respectively.

---

# What is the average daily activity pattern?

```{r avg_daily_activity}
tabdata2<- rawdata %>%
  group_by(interval) %>%
  summarize(mean_steps=mean(steps, na.rm=TRUE))

plot(tabdata2$interval, tabdata2$mean_steps, pch=2, type="l", main="Average Number of Steps Taken (averaged across all days)", ylab="Average Number of Steps", xlab="Five-Minute Interval")
```

The 5-minute interval, on average across all the days in the dataset, that contains the maximum number of steps is interval **`r tabdata2$interval[which.max(tabdata2$mean_steps)]`**.

---
# Imputing missing values

The total number of missing values in the dataset (i.e. the total number of rows with NAs) is **`r sum(is.na(rawdata$steps))`**.

Missing values will be imputed using the median across all days for an interval


```{r data_impute}
  tabdata3 <- rawdata %>%
  group_by(interval) %>%
  summarize(median_steps=median(steps, na.rm=TRUE))
  rawdata_merge <- merge(rawdata,tabdata3, by="interval", sort=FALSE)
  rawdata_merge <- arrange(rawdata_merge, date,interval)
  indices_na <- which(is.na(rawdata$steps)==TRUE)
  rawdata_nona <- rawdata_merge
  rawdata_nona[indices_na,"steps"] <- rawdata_merge[indices_na,"median_steps"]
  tabdata4 <- rawdata_nona %>%
    group_by(date) %>%
    summarize(Total_Steps=sum(steps,na.rm=TRUE))
  hist(tabdata4$Total_Steps, main="Total Steps Taken Per Day", xlab="Number of Steps", ylab="Frequency")  
```


The mean and median number of steps taken per day were **`r format(mean(tabdata4$Total_Steps, na.rm=TRUE),scientific=FALSE)`** and **`r median(tabdata4$Total_Steps, na.rm=TRUE)`**, respectively. These values differ from the previously calculated estimates. Both of these values for the steps per day are higher than the values computed for the data without missing values imputed.


---

# Are there differences in activity patterns between weekdays and weekends?

```{r weekday_weekend_anal}
  rawdata_nona$weekday <- weekdays(rawdata_nona$date)
  rawdata_nona$weekday_fac <- factor((rawdata_nona$weekday %in% c("Saturday","Sunday")), levels=c("TRUE","FALSE"),labels=c("weekend","weekday"))
  
  tabdata5 <- rawdata_nona %>%
    select(-median_steps) %>%
    group_by(weekday_fac, interval) %>%
    summarize(mean_steps=mean(steps, na.rm=TRUE))

par(mfrow=c(2,1), mar=c(4, 4, 2, 2))
with(subset(tabdata5,tabdata5$weekday_fac=="weekend"),
plot(interval,mean_steps, pch=2, type="l", main="Average Number of Steps Taken by Day of Week",ylab="Average Number of Steps", xlab="Five-Minute Interval"))
legend(50,125,legend="Weekend")
with(subset(tabdata5,tabdata5$weekday_fac=="weekday"),
plot(interval,mean_steps, pch=2, type="l", ylab="Average Number of Steps", xlab="Five-Minute Interval"))
legend(20,150,legend="Weekday")
```
 
Overall, the number of steps taken during the weekend are higher than on weekdays.
